class: inverse, center, middle
# {gtsummary} review

---
# Summarizing many variables

.medium[
Often will need a "Table 1" describing many characteristics of a study population either **overall**, or **according to another variable**. 

There are many packages to do this in R, including the {gtsummary} package.

Install and load the {gtsummary} package:

```{r eval = FALSE}
install.packages("gtsummary")
```

```{r}
library(gtsummary)
```

The summary statistics can be customized in many ways. See [the package website](https://www.danieldsjoberg.com/gtsummary/index.html) for more details and many tutorials.

Note that the functions from the {gtsummary} package are **particularly useful** when used **in conjunction with R Markdown**, since they are meant to produce publication-ready tables.
]

---
# Overall summary tables

.medium[
Make **overall descriptive statistic tables** using the `tbl_summary()` function. Subset the data for the variables of interest within the function:
]

.pull-left[
```{r ex-tbl, eval = FALSE}
tbl_summary(
  data = trial[, c("marker", "grade")]
  )
```
]

.pull-right[
```{r ref.label="ex-tbl", echo = FALSE}
```
]


---
# Cross-tabulated summary tables

.medium[
Make tables of **descriptive statistics according to another variable** using the `by` argument to the `tbl_summary()` function:
]

.pull-left[
```{r ex-tbl-by, eval = FALSE}
tbl_summary(
  trial[, c("marker", "grade", "trt")], 
  by = "trt"
)
```
]

.pull-right[
```{r ref.label = "ex-tbl-by", echo = FALSE}
```
]


