class: inverse, center, middle
# Bivariable statistical tests

---
# Two categorical variables

.medium[

There are two main tests for associations between **two categorical variables**: the **chi-squared** test and **Fisher's exact** test. 

Conduct these tests in R using the following functions:

- `chisq.test()` 
- `fisher.test()`

]


---
# Chi-squared test

.medium[
Conduct a chi-squared test of the **null hypothesis that there is no association** between treatment and response versus the alternative hypothesis that there is an association between treatment and response.

```{r}
chisq.test(x = trial$trt, y = trial$response)
```

The chi-squared test statistic is `r round(chisq.test(trial$trt, trial$response)$statistic, 5)` with associated p-value `r round(chisq.test(trial$trt, trial$response)$p.value, 4)`. 

**Do not reject the null hypothesis** since the p-value is greater than the traditional threshold for significance of 0.05. 
]

---
# Fisher's exact test

.medium[
Conduct a Fisher's exact test (an alternative to the chi-squared test when any **expected cell count is <5**):

```{r}
fisher.test(x = trial$trt, y = trial$response)
```

The p-value is `r round(fisher.test(trial$trt, trial$response)$p.value, 4)`.

Do not reject the null hypothesis of no association between treatment and response since the p-value is greater than the traditional threshold for significance of 0.05.  
]

---
# An R programming aside - the pipe

.large[
The **pipe** operator: `%>%`

- Strings multiple functions together into a sequence
- From the {magrittr} package but loaded with many other packages including {gtsummary}
- Allows for code that is easier to read and understand
- Replaces nested functions
]


---
# First, create a basic contingency table

.medium[
Use `tbl_summary()` to create a table of tumor grade according to response:
]

.pull-left[
```{r basic-tbl, eval = FALSE}
tbl_summary(
  data = trial[, c("grade", "response")],
  by = response
  )
```
]

.pull-right[
```{r ref.label = "basic-tbl", echo = FALSE}
```
]

<br>
<br>

.medium[
*Note the message that missing values in response have been removed from this table and offering an option to include them as an additional column*
]


---
# Next, add p-values

.large[
Use the function `add_p()` to add p-values to this table. 

By default, this function will use the chi-squared test *without* the continuity correction if all expected cell counts are $\geq5$ and will use Fisher's exact test if any expected cell count is <5. 

The primary argument to `add_p()` is the results of a call to `tbl_summary()`.

But **how** do we pass the table results to this function?
]


---
# Approach 1: Intermediate object

.medium[
Create an intermediate object "mytbl" for the results of `tbl_summary()`:
]

.pull-left[
```{r app1, eval = FALSE, message = FALSE}
mytbl <- tbl_summary(
  data = trial[, c("grade", "response")],
  by = response
  )

add_p(mytbl)
```
]

.pull-right[
```{r ref.label = "app1", echo = FALSE}
```
]

<br>
<br>

.medium[
The **disadvantage** of this approach is the creation and storage of an object  **not needed again**.
]


---
# Approach 2: Nested functions

.medium[
Wrap `tbl_summary()` inside `add_p()`:
]

.pull-left[
```{r app2, eval = FALSE, message = FALSE}
add_p(
  tbl_summary(
    data = trial[, c("grade", "response")],
    by = response
    )
  )
```
]

.pull-right[
```{r ref.label="app2", echo = FALSE}
```
]

<br>
<br>

.medium[
The **disadvantage** is that this code can be **hard to read** and it is read **from the inside out**.
] 


---
# Approach 3: The pipe operator

.medium[
Use the pipe operator to string the two functions together:
]

.pull-left[
```{r "app3", eval = FALSE, message = FALSE}
tbl_summary(
  data = trial[, c("grade", "response")],
  by = response
  ) %>% 
  add_p()
```
]

.pull-right[
```{r ref.label="app3", echo = FALSE}
```
]

<br>
<br>

.medium[
The **left hand side** is passed as the **first argument** to the function on the **right hand side**.
]


---
# One continuous, one categorical

.medium[
The most common statistical tests for the association between a continuous and a categorical variable are the **non-parametric Kruskal-Wallis** test and the **parametric t-test**. 

Conduct these tests in R using the following functions:

- `kruskal.test()` (see also `wilcox.test()`)
- `t.test()`

*Note that the Kruskal-Wallis test is also known as the **Wilcoxon rank-sum** test in the special case of a 2-level categorical variable.*
]


---
# Kruskal-Wallis test

.medium[

Conduct a Kruskal-Wallis test to test the **null hypothesis of no association between marker status and response** versus the alternative hypothesis that there is an association between marker status and response.

```{r}
kruskal.test(marker ~ response, data = trial)
```

]


---
# Wilcoxon rank-sum test

.medium[

Since "response" is a 2-level categorical variable the Wilcoxon rank-sum test (without continuity correction) produces the **same result**:

```{r}
wilcox.test(marker ~ response, correct = FALSE, data = trial)
```

With a p-value of `r round(wilcox.test(marker ~ response, correct = FALSE, data = trial)$p.value, 3)`, **do not reject** the null hypothesis at the 0.05 significance level.
]


---
# t-test

.medium[
If appropriate given the distribution of the continuous variable, conduct a t-test:

```{r}
t.test(marker ~ response, data = trial)
```

The **p-value differs** but the **conclusion is the same**: do not reject the null hypothesis that there is no association between marker and response.
]


---
# Histogram of marker distribution

.large[

.pull-left[
.medium[
In this example the data are **right-skewed** so a t-test is **not appropriate**.

```{r echo = FALSE}
library(ggplot2)
```

```{r hist-code, eval = FALSE}
ggplot(data = trial, aes(x = marker)) +
  geom_histogram(bins = 10)
```
]
]

.pull-right[

```{r ref.label = "hist-code", echo = FALSE}
```

]

]


---
# Multiple continuous variables

.medium[
Use `tbl_summary()` with `add_p()` from the {gtsummary} package to summarize **multiple continuous variables** with respect to a categorical variable:
]

.pull-left[
```{r mult-cont, eval = FALSE, message = FALSE}
trial[, c("age", "marker", "response")] %>% 
  tbl_summary(
    by = response
    ) %>% 
  add_p()
```
]

.pull-right[
```{r ref.label = "mult-cont", echo = FALSE}
```
]



---
# Mixture of variable types

.pull-left[
```{r mix-tbl, eval = FALSE, message = FALSE}
trial[, c("grade", "marker", "response")] %>% 
  tbl_summary(
    by = response
    ) %>% 
  add_p()
```
]

.pull-right[
```{r ref.label = "mix-tbl", echo = FALSE}
```
]

