class: inverse, center, middle
# Adjustment for multiple testing


---
# Example of many p-values

.medium[

.pull-left[
The **false-discovery rate** correction and **other approaches** can be implemented with the R function `p.adjust()`. 

P-values for association between 10 gene mutations and treatment response:
]


.pull-right[
```{r echo = FALSE}
tibble::tibble(
  Gene = paste0("gene", seq(1:10)),
  `p-value` = c(0.001, 0.245, 0.784, 0.034, 0.004, 0.123, 0.089, 0.063, 0.228, 
                0.802)
  ) %>% 
  knitr::kable()
```

]

]


---
# False-discovery rate adjustment

.medium[
Adjust these for multiple testing using the false-discovery rate approach. Pass the vector of p-values to `p.adjust()` and specify `method = "fdr"`:

```{r}
p.adjust(
  p = c(0.001, 0.245, 0.784, 0.034, 0.004, 0.123, 0.089, 0.063, 0.228, 0.802),
  method = "fdr"
)
```

Get back a **vector of the adjusted p-values**, listed in the same order as the originally provided p-values.
]



---
# Using the {gtsummary} package

.medium[
Add adjusted p-values to results tables made with the {gtsummary} package using the function `add_q()`:
]

.pull-left[
```{r mult-comp, eval = FALSE, message = FALSE}
tbl_summary(
  trial[, c("age", "marker", "trt")],
  by = trt
  ) %>% 
  add_p() %>% 
  add_q()
```

*Note that "q-value" is a common term for p-values that have been adjusted for multiple comparisons*.

By default, the FDR correction is applied, but all methods from `p.adjust()` can be specified using the `method` argument to `add_q()`.
]

.pull-right[
```{r ref.label = "mult-comp", echo = FALSE}
```

*Note that this table is for demonstration purposes only*.
]



