class: inverse, center, middle
# Basic power calculations


---
# Get help!

.xlarge[
Most importantly: **consult a statistician**
]

<br>

.large[
Opinion: R is **not the best** for power and sample size calculations. Commercial software programs like **EAST** and **PASS** are easier to use and have more options.
]


---
# R functions for power calculations

.large[
But for the most **basic power calculations**, the following functions are available in R:

- `power.t.test()`
- `power.prop.test()`
]



---
# Power for a t-test

.medium[
`power.t.test()` can compute the power of a **one-sample** or **two-sample** t-test. The main arguments are:

`n`: sample size

`delta`: true difference in means

`sd`: standard deviation

`sig.level`: type I error probability, i.e. alpha

`power`: power, i.e. 1 minus type II error probability

<br>

Exactly **one** of these arguments can be specified as `NULL` and that parameter will be estimated based on the others. 
]


---
# Example of t-test power

.medium[
What **sample size** is needed to detect a difference of **0.5 units** in mean biomarker level between two treatment groups when the **standard deviation is 0.8**, the **type I error is 0.05** and the **type II error is 0.1**.
]

.pull-left[
```{r tpower, eval = FALSE}
power.t.test(n = NULL, 
             delta = 0.5, 
             sd = 0.8,
             sig.level = 0.05,
             power = 0.9, 
             type = "two.sample",  
             alternative = "one.sided")
```
]

.pull-right[
```{r ref.label="tpower", echo = FALSE}
```
]

We find that we need **45 patients in each group** in order to conduct this test.

---
# Power for a proportion test

.medium[
`power.prop.test()` can compute the power of the **two-sample test for proportions**. The main arguments are:

`n`: sample size

`p1`: probability in group 1

`p2`: probability in group 2

`sig.level`: type I error probability, i.e. alpha

`power`: power, i.e. 1 minus type II error probability 

<br>

Exactly **one** of these arguments can be specified as `NULL` and that parameter will be estimated based on the others. 
]


---
# Example of proportion test power

.medium[
What **power** do we have to detect a difference between a **0.34 response rate** and a **0.54 response rate** if we have **100 patients in each group** at a **type I error rate of 0.05**.
]

.left-column[
```{r proppower, eval = FALSE}
power.prop.test(
  n = 100,
  p1 = 0.34,
  p2 = 0.54,
  sig.level = 0.05,
  power = NULL,
  alternative = "one.sided"
)
```
]

.right-column[
```{r ref.label = "proppower", echo = FALSE}
```
]

<br>

And we find that we have **89% power** for this test.

